maze-dataset: Maze Generation with Algorithmic Variety and Representational Flexibility

### 2025-04-26 docs/paper/paper.tex
cut down impl and sre sections, better links

\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{calc}
\tikzset{ % Define a TikZ style for an external hyperlink node.
  hyperlink node url/.style={
    alias=sourcenode,
    append after command={
      let \p1 = (sourcenode.north west),
          \p2 = (sourcenode.south east),
          \n1 = {\x2-\x1},
          \n2 = {\y1-\y2} in
      node[inner sep=0pt, outer sep=0pt, anchor=north west, at=(\p1)]
          {\href{#1}{\XeTeXLinkBox{\phantom{\rule{\n1}{\n2}}}}}
    }
  }
}
\providecommand{\XeTeXLinkBox}[1]{#1}
\newcommand{\docslink}[2]{\href{https://understanding-search.github.io/maze-dataset/#1}{#2}}
\newcommand{\docslinkcode}[2]{\href{https://understanding-search.github.io/maze-dataset/#1}{\texttt{#2}}}
\newcommand{\secref}[1]{\hyperref[#1]{section: \textit{\nameref{#1}}}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\begin{figure}
\hypertarget{fig:benchmarks}{%
\centering
\includegraphics[width=0.9\textwidth,height=\textheight]{figures/benchmarks/gridsize-vs-gentime.pdf}
\caption{Plot of maze generation time. Generation time scales
exponentially with maze size for all algorithms. Generation time per
maze does not depend on the number of mazes being generated, and there
is minimal overhead to initializing the generation process for a small
dataset. Wilson's algorithm is notably less efficient than others and
has high variance. Note that values are averaged across all parameter
sets for that algorithm. More information can be found on the
\href{https://understanding-search.github.io/maze-dataset/benchmarks/}{benchmarks
page}.}\label{fig:benchmarks}
}
\end{figure}

@@

In order to replicate the exact dataset distribution of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}), the parameter
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#MazeDatasetConfig.endpoint_kwargs}{\texttt{MazeDatasetConfig.endpoint\_kwargs:}}
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#EndpointKwargsType}{\texttt{EndpointKwargsType}}
allows for additional constraints such as enforcing that the start or
end point be in a ``dead end'' with only one accessible neighbor cell.
However, combining these constraints with cyclic mazes (such as those
generated with percolation), as was required for the work in
(\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et al.,
2024}), can lead to an absence of valid start and end points. Placing
theoretical bounds on this success rate is difficult, as it depends on
the exact maze generation algorithm and parameters used. To deal with
this, our package provides a way to estimate the success rate of a given
configuration using a symbolic regression model trained with PySR
(\protect\hyperlink{ref-pysr}{Cranmer, 2023}). More details on this can
be found in
\href{https://understanding-search.github.io/maze-dataset/notebooks/estimate_dataset_fractions.html}{\texttt{estimate\_dataset\_fractions.ipynb}}.
Using the estimation algorithm simply requires the user to call
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDatasetConfig.success_fraction_compensate}{\texttt{cfg\_new:\ MazeDatasetConfig\ =\ cfg.success\_fraction\_compensate()}},
providing their initial \texttt{cfg} and then using the returned
\texttt{cfg\_new} in its place.

\hypertarget{success-rate-estimation-algorithm}{%
\subsubsection{Success Rate Estimation
Algorithm}\label{success-rate-estimation-algorithm}}

The base function learned by symbolic regression provides limited
insight and may be subject to change. It is defined as
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/success_predict_math.html\#cfg_success_predict_fn}{\texttt{cfg\_success\_predict\_fn}},
and takes a 5 dimensional float vector created by
\texttt{MazeDatasetConfig.\_to\_ps\_array()} which represents the
{[}percolation value, grid size, endpoint deadend configuration,
endpoint uniqueness, categorical generation function index{]}.

However, the outputs of this function are not directly usable due to
minor divergences at the endpoints with respect to the percolation
probability \(p\). Since we know that maze success is either guaranteed
or impossible for \(p=0\) and \(p=1\), we define the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/success_predict_math.html\#soft_step}{\texttt{soft\_step}}
function to nudge the raw output of the symbolic regression. This
function is defined with the following components:

shifted sigmoid \(\sigma_s\), amplitude scaling \(A\), and \(h\)
function given by \[
  \sigma_s(x) = (1 + e^{-10^3 \cdot (x-0.5)})^{-1}
  \qquad A(q,a,w) = w \cdot (1 - |2q-1|^a)
\] \[
  h(q,a) = q \cdot (1 - |2q-1|^a) \cdot (1-\sigma_s(q)) + (1-(1-q) \cdot (1 - |2(1-q)-1|^a)) \cdot \sigma_s(q)
\]

We combine these to get the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/success_predict_math.html\#soft_step}{\texttt{soft\_step}}
function, which is identity-like for \(p \approx 0.5\), and pushes \(x\)
to extremes otherwise. \[
  \text{soft\_step}(x, p, \alpha, w) = h(x, A(p, \alpha, w))
\]

Finally, we define \[
  \text{cfg\_success\_predict\_fn}(\mathbf{x}) = \text{soft\_step}(\text{raw\_val}, x_0, 5, 10)
\]

where \texttt{raw\_val} is the output of the symbolic regression model.
The parameter \(x_0\) is the percolation probability, while all other
parameters from \texttt{\_to\_ps\_array()} only affect
\texttt{raw\_val}.

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=\textheight]{figures/ep/ep_deadends_unique-crop.pdf}
\caption{An example of both empirical and predicted success rates as a
function of the percolation probability \(p\) for various maze sizes,
percolation with and without depth first search, and
\texttt{endpoint\_kwargs} requiring that both the start and end be in
unique dead ends. Empirical measures derived from a sample of 128 mazes.
More information can be found on the
\href{https://understanding-search.github.io/maze-dataset/benchmarks/}{benchmarks
page}.}
\end{figure}

@@

We refer to our
\href{https://github.com/understanding-search/maze-dataset}{repository}
and
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html}{docs}
for documentation and up-to-date implementation details.

This package utilizes a simple, efficient representation of mazes as
subgraphs of a finite lattice, which we call a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze}{\texttt{LatticeMaze}}.
Using an adjacency matrix for storing mazes would be memory inefficient
by failing to exploit the highly sparse structure -- for example, for a
2-dimensional maze, only 4 off-diagonal bands would be have nonzero
values. On the other hand, using an adjacency list could lead to a poor
lookup time for whether any given connection exists.

Instead, we describe mazes with the following representation: for a
\(2\)-dimensional lattice with \(r\) rows and \(c\) columns, we
initialize a boolean array \[
  A = \{0, 1\}^{2 \times r \times c}
\] which we refer to in the code as a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.connection_list}{\texttt{connection\_list}}.
The value at \(A[0,i,j]\) determines whether a \emph{downward}
connection exists from node \([i,j]\) to \([i+1, j]\). Likewise, the
value at \(A[1,i,j]\) determines whether a \emph{rightward} connection
to \([i, j+1]\) exists. Thus, we avoid duplication of data about the
existence of connections and facilitate fast lookup time, at the cost of
requiring additional care with indexing. Note that this setup allows for
a periodic lattice. Generation of mazes is detailed in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMazeGenerators}{\texttt{LatticeMazeGenerators}}.

To produce solutions to mazes, two points are selected uniformly at
random without replacement from the connected component of the maze, and
the \(A^*\) algorithm (\protect\hyperlink{ref-A_star}{Hart et al.,
1968}) is applied to find the shortest path between them. The endpoint
selection can be controlled via
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#MazeDatasetConfig.endpoint_kwargs}{\texttt{MazeDatasetConfig.endpoint\_kwargs:}}
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#EndpointKwargsType}{\texttt{EndpointKwargsType}},
and complications caused by this are detailed in
\hyperref[sec:success-rate-estimation]{section: \textit{\nameref{sec:success-rate-estimation}}}.
A maze with a solution is denoted a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#SolvedMaze}{\texttt{SolvedMaze}},
which inherits from
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze}{\texttt{LatticeMaze}}.

Parallelization is implemented via the \texttt{multiprocessing} module
in the Python standard library, and parallel generation can be
controlled via keyword arguments to
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.from_config}{\texttt{MazeDataset.from\_config()}}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-A_star}{}}%
Hart, P. E., Nilsson, N. J., \& Raphael, B. (1968). A {Formal Basis} for
the {Heuristic Determination} of {Minimum Cost Paths}. \emph{IEEE
Transactions on Systems Science and Cybernetics}, \emph{4}(2), 100--107.
\url{https://doi.org/10.1109/TSSC.1968.300136}

>>>>

\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{calc}
\tikzset{ % Define a TikZ style for an external hyperlink node.
  hyperlink node url/.style={
    alias=sourcenode,
    append after command={
      let \p1 = (sourcenode.north west),
          \p2 = (sourcenode.south east),
          \n1 = {\x2-\x1},
          \n2 = {\y1-\y2} in
      node[inner sep=0pt, outer sep=0pt, anchor=north west, at=(\p1)]
          {\href{#1}{\XeTeXLinkBox{\phantom{\rule{\n1}{\n2}}}}}
    }
  }
}
\providecommand{\XeTeXLinkBox}[1]{#1}
\newcommand{\docslink}[2]{\href{https://understanding-search.github.io/maze-dataset/#1}{#2}}
\newcommand{\docslinkcode}[2]{\href{https://understanding-search.github.io/maze-dataset/#1}{\texttt{#2}}}
\newcommand{\docslinkcodemain}[1]{\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\##1}{\texttt{#1}}}
\newcommand{\secref}[1]{\hyperref[#1]{section: \textit{\nameref{#1}}}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\input{figures/tex/fig6_benchmarks.tex}

@@

In order to replicate the exact dataset distribution of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}), the parameter
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#MazeDatasetConfig.endpoint_kwargs}{\texttt{MazeDatasetConfig.endpoint\_kwargs:}}
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#EndpointKwargsType}{\texttt{EndpointKwargsType}}
allows for additional constraints such as enforcing that the start or
end point be in a ``dead end'' with only one accessible neighbor cell.
However, combining these constraints with cyclic mazes (such as those
generated with percolation), as was required for the work in
(\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et al.,
2024}), can lead to an absence of valid start and end points. Placing
theoretical bounds on this success rate is difficult, as it depends on
the exact maze generation algorithm and parameters used. To deal with
this, our package provides a way to estimate the success rate of a given
configuration using a symbolic regression model trained with PySR
(\protect\hyperlink{ref-pysr}{Cranmer, 2023}). More details on this can
be found in
\href{https://understanding-search.github.io/maze-dataset/notebooks/estimate_dataset_fractions.html}{\texttt{estimate\_dataset\_fractions.ipynb}}.

\input{figures/tex/fig7_sre.tex}

@@

Using an adjacency matrix for storing mazes would be memory inefficient
by failing to exploit the highly sparse structure, while using an
adjacency list could lead to a poor lookup time. This package utilizes a
simple, efficient representation of mazes as subgraphs of a finite
lattice, which we call a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze}{\texttt{LatticeMaze}}.

We describe mazes with the following representation: for a
\(2\)-dimensional lattice with \(r\) rows and \(c\) columns, we
initialize a boolean array \[
  A = \{0, 1\}^{2 \times r \times c}
\] which we refer to in the code as a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.connection_list}{\texttt{connection\_list}}.
The value at \(A[0,i,j]\) determines whether a \emph{downward}
connection exists from node \([i,j]\) to \([i+1, j]\). Likewise, the
value at \(A[1,i,j]\) determines whether a \emph{rightward} connection
to \([i, j+1]\) exists. Thus, we avoid duplication of data about the
existence of connections and facilitate fast lookup time, at the cost of
requiring additional care with indexing. Note that this setup allows for
a periodic lattice. Generation of mazes is detailed in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMazeGenerators}{\texttt{LatticeMazeGenerators}}.

@@
